---
title: "KENYA'S POPULATION GROWTH 1950 - 2020"
author: "Karanja David"
date: "16/12/2021"
output: github_document
---

# ANALYSIS OF POPULATION GROWTH IN KENYA 

## setting up the enviroment
loading packages
```{r setting up the enviroment, message=FALSE, warning=FALSE}
library(animation)
library(gganimate)
library(janitor)
library(lubridate)
library(tidyverse)
library(wpp2019)
```


## load data sets
```{r data}
#load data from wpp2019 package
data(popF)
data(popM)

```


## explore data sets
```{r data exploration}
#popF contains female population while popM contains male population.
str(popF)
str(popM)
```


## Transform the data

### Kenya's female population

Country names are derived from country codes specified in the link [codes](https://en.wikipedia.org/wiki/ISO_3166-1_numeric)

```{r transforming female population data }
# Country of interest - Kenya code 404.

#Kenya's Female population
pop_female_kenya <- popF %>%
  gather(Year, popF, '1950'|'1955'|'1960'|'1965'|'1970'|'1975'
         |'1980'|'1985'|'1990'|'1995'|'2000'|'2005'
         |'2010'|'2015'|'2020') %>%
  clean_names() %>% 
  rename(pop_in_thousands = pop_f, country = name) %>% 
  filter(country_code %in% c(404)) %>%
  select(country, age, year, pop_in_thousands) 

#add a gender column
pop_female_kenya$gender <- "female"

#view a sample of the new data frame
head(pop_female_kenya)



```

### Kenya's male population

```{r transforming male population data}
#Kenya's Male population
pop_male_kenya <- popM %>%
  gather(Year, popM, '1950'|'1955'|'1960'|'1965'|'1970'|'1975'
         |'1980'|'1985'|'1990'|'1995'|'2000'|'2005'
         |'2010'|'2015'|'2020') %>%
  clean_names() %>% 
  rename(pop_in_thousands = pop_m, country = name) %>% 
  filter(country_code %in% c(404)) %>%
  select(country, age, year, pop_in_thousands) 

#add a gender column
pop_male_kenya$gender <- "male"

#view a sample of the new data frame
head(pop_male_kenya)

```


### Combine male and female data frames 
```{r row binding data frames}
#ensure consistency of column names and corresponding data types
#combine male and female population
total_pop <- rbind(pop_female_kenya, pop_male_kenya)
```


## Clean the data types
### 1. age: character to factor
```{r convert age column data type}
#1) age to factor
total_pop$age <- factor(total_pop$age, levels = c("0-4", "5-9", "10-14", 
                                                  "15-19", "20-24", "25-29",
                                                  "30-34", "35-39", "40-44",
                                                  "45-49", "50-54", "55-59",
                                                  "60-64", "65-69", "70-74",
                                                  "75-79", "80-84", "85-89",
                                                  "90-94", "95-99", "100+"),
                        order = TRUE)
```


### 2. gender: character to factor
```{r convert gender column data type}

#2)gender to factor
total_pop$gender <- factor(total_pop$gender) 
```


### 3. year: character to date to numeric
```{r convert year column data type}
#3) year to numeric
total_pop$year <- as.Date(total_pop$year, "%Y")

total_pop$year <- year(total_pop$year)
```


### View structure of the transformed total_pop data frame
```{r structure}
#view the correct data types 
str(total_pop)
```

## plot Kenya's 2020 population distribution
```{r 2020 data plot}
#Prepare the data

kenya_population <- total_pop %>%
  mutate(pop = pop_in_thousands * 1000,
         population = ifelse(gender == "female", 1, -1)*pop)


#plot Kenya's population distribution between male and female in 2020

kenya_population_plot <- kenya_population %>% 
  filter(year == 2020) %>% 
  ggplot(aes(x = age, y = population, fill = gender))+ 
  geom_bar(stat="identity") +
  scale_y_continuous(breaks=seq(-4000000, 4000000, 1000000),
                     labels =c('4 Mil','3 Mil','2 Mil', '1 Mil',
                               '0','1 Mil','2 Mil','3 Mil', '4 Mil'),
                     limits = c(-4000000, 4000000)) +
  xlab("Age Groups") +
  coord_flip() +
  
  labs(title ='Population in Kenya in 2020') +
  scale_fill_manual(values=c('deeppink1', 'navy blue'))+
  theme(plot.background = element_rect(fill = "limegreen", 
                                       color = NA),
        panel.background = element_rect(fill = "limegreen"),
        axis.text = element_text(color = "white"),
        axis.title = element_text(color = "white", size = 12),
        plot.title = element_text(color = "white"),
        # plot.margin = unit(c(2,1.5,1,1), "cm"),
        panel.grid = element_line(color = "#4d4d4d"))

  
 kenya_population_plot
```


## gif showing population distribution since 1950

```{r gif}
#visualize the data over the years in a gif
 
 kenya_population_gif <- ggplot(data = kenya_population,
                 aes(x = age, y = population, fill = gender))+
   geom_bar(stat='identity') + 
   scale_y_continuous(breaks=seq(-4000000, 4000000, 1000000),
                      labels =c('4 Mil','3 Mil','2 Mil', '1 Mil',
                                '0','1 Mil','2 Mil','3 Mil', '4 Mil'),
                      limits = c(-4000000, 4000000)) +
   xlab("Age Groups") + 
   ylab("Population") +  
   coord_flip() + 
   labs(fill = "gender") +
   scale_fill_manual(values=c('deeppink1', 'blue2')) +
   #theme_bw()+
   theme(plot.background = element_rect(fill = "white", color = NA),
         panel.background = element_rect(fill = "chartreuse"),
         axis.text = element_text(color = "black"),
         axis.title = element_text(color = "black", size = 12),
         plot.title = element_text(color = "black"),
         panel.grid = element_line(color = "grey80")) + 
   transition_time(year) +
   labs(title = "{as.integer(frame_time)} Population in Kenya")

 
kenya_population_gif
 
```




